---
title: "Differential miRNA Expression Analysis - TCGA"
output: html_notebook
---

## I. Preliminaries

### Loading libraries

```{r}
library("tidyverse")
library("tibble")
library("msigdbr")
library("ggplot2")
library("TCGAbiolinks")
library("RNAseqQC")
library("DESeq2")
library("ensembldb")
library("purrr")
library("magrittr")
library("vsn")
library("matrixStats")
library("dplyr")
library("grex")
library("biomaRt")
```

## II. Downloading TCGA miRNA expression data 

Download miRNA expression data from The Cancer Genome Atlas (TCGA):
- `TCGA-COAD` refers to the biospecimen data for colon adenocarcinoma.

```{r, message=FALSE}
query_tumor <- GDCquery(
  project = "TCGA-COAD",
  data.category = "Transcriptome Profiling",
  data.type = "miRNA Expression Quantification",
  experimental.strategy = "miRNA-Seq",
  access = "open",
  sample.type = "Primary Tumor"
)
tumor <- getResults(query_tumor)
tumor
```

```{r, message=FALSE}
query_normal <- GDCquery(
  project = "TCGA-COAD",
  data.category = "Transcriptome Profiling",
  data.type = "miRNA Expression Quantification",
  experimental.strategy = "miRNA-Seq",
  access = "open",
  sample.type = "Solid Tissue Normal"
)
normal <- getResults(query_normal)
normal
```
Consider only samples with both normal and malignant tissues.

```{r, message=FALSE}
submitter_ids <- inner_join(tumor, normal, by = "cases.submitter_id") %>%
  dplyr::select(cases.submitter_id)
tumor <- tumor %>%
  dplyr::filter(cases.submitter_id %in% submitter_ids$cases.submitter_id)
normal <- normal %>%
  dplyr::filter(cases.submitter_id %in% submitter_ids$cases.submitter_id)

samples <- rbind(tumor, normal)
invisible(unique(samples$sample_type))
samples
```

Download only samples with both normal and malignant tissues.

To impose this filtering, we set the `barcode` argument of `GDCquery` to `samples$sample.submitter_id` (which was generated in the previous cell).

```{r, message=FALSE}
query_coad <- GDCquery(
  project = "TCGA-COAD",
  data.category = "Transcriptome Profiling",
  data.type = "miRNA Expression Quantification",
  experimental.strategy = "miRNA-Seq",
  access = "open",
  sample.type = c("Solid Tissue Normal", "Primary Tumor"),
  barcode = as.list(samples$sample.submitter_id)
)
```
If this is your first time running this notebook (i.e., you have not yet downloaded the results of the query in the previous block), uncomment the code block below.

```{r, message=FALSE}
# GDCdownload(query_coad)
```

Running the code block above should generate and populate a directory named `GDCdata`.

## III. Data preprocessing

Construct the RNA-seq count matrix.

```{r, echo = TRUE, results="hide"}
tcga_coad_data <- GDCprepare(query_coad, summarizedExperiment = TRUE)
```

```{r}
rownames(tcga_coad_data) <- tcga_coad_data$miRNA_ID
count_matrix <- tcga_coad_data[, colnames(tcga_coad_data)[grep("count", colnames(tcga_coad_data))]]
colnames(count_matrix) <- gsub("read_count_", "", colnames(count_matrix))

# Remove duplicate entries
count_matrix_df <- data.frame(count_matrix)
count_matrix_df <- count_matrix_df[!duplicated(count_matrix_df), ]
count_matrix <- data.matrix(count_matrix_df)
rownames(count_matrix) <- cleanid(rownames(count_matrix))
count_matrix <- count_matrix[!(duplicated(rownames(count_matrix)) | duplicated(rownames(count_matrix), fromLast = TRUE)), ]

head(count_matrix[1:5, 1:4])
```
Format the `samples` table so that it can be fed as input to DESeq2.

```{r}
rownames(samples) <- samples$cases
samples <- samples %>%
  dplyr::select(case = "cases.submitter_id", type = "sample_type")
samples$type <- str_replace(samples$type, "Solid Tissue Normal", "normal")
samples$type <- str_replace(samples$type, "Primary Tumor", "tumor")
```

DESeq2 requires the row names of `samples` should be identical to the column names of `count_matrix`.

```{r, echo = TRUE, results="hide"}
colnames(count_matrix) <- gsub(x = colnames(count_matrix), pattern = "\\.", replacement = "-")
count_matrix <- count_matrix[, rownames(samples)]

# Sanity check
all(colnames(count_matrix) == rownames(samples))
```

## IV. Differential miRNA expression analysis

References: 

- Official documentation: https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html
- Good balance of theory and hands-on: https://hbctraining.github.io/DGE_workshop/lessons/04_DGE_DESeq2_analysis.html
- Quality control: https://cran.r-project.org/web/packages/RNAseqQC/vignettes/introduction.html

Construct the `DESeqDataSet` object.

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = samples,
  design = ~type
)
```

### Quality Control

Display quality control (QC) plots (refer to https://cran.r-project.org/web/packages/RNAseqQC/vignettes/introduction.html)

- Total sample counts  
  - Total number of counts for each sample
  - We typically expect all samples to have total counts within the same order of magnitude
  
- Library complexity
  - What fraction of counts is taken up by what fraction of genes
  - Samples showing a different library complexity than the rest might be considered low quality
  
- Gene detection
  - Number of detected genes for each sample
  
**CAVEAT:** There seem to be some outliers — but we defer handling them for now!

```{r, message=FALSE}
plot_total_counts(dds)
plot_library_complexity(dds)
plot_gene_detection(dds)
```
Perform miRNA filtering.

We determined `min_count` empirically by looking at the red trend line in the variance stabilization plot. Ideally, this trend line should be flat (i.e., stable).

```{r}
dds <- filter_genes(dds, min_count = 10)
```

Transform the read counts.

From https://chipster.csc.fi/manual/deseq2-transform.html: <br>
You can use the resulting transformed values only for visualization and clustering, not for differential expression analysis which needs raw counts.

```{r}
dds <- estimateSizeFactors(dds)
nsub <- sum(rowMeans(counts(dds, normalized = TRUE)) > 10)
vsd <- vst(dds, nsub = nsub)
mean_sd_plot(vsd)
```

Check the clustering of the samples. 

If you encounter the error `Error in loadNamespace(x) : there is no package called 'ComplexHeatmap'`, uncomment and run the following code block:

```{r}
# install.packages("devtools", dependencies = TRUE)
# devtools::install_github("jokergoo/ComplexHeatmap")
```

```{r, fig.width=10,fig.height=15}
set.seed(42)
plot_sample_clustering(vsd, anno_vars = c("type"), distance = "euclidean")
```

Perform principal component analysis (PCA).

```{r}
plot_pca(vsd, PC_x = 1, PC_y = 2, shape_by = "type")
```
### Regulated Cell Death

Refer to `1. Exploratory Data Analysis - MSigDB Gene Sets + GTEx TPM.rmd` for more detailed documentation on obtaining the gene sets.

#### Mapping miRNAs to target mRNAs

We are going to refer to miRDB to for the miRNA-mRNA mapping: https://academic.oup.com/nar/article/48/D1/D127/5557729

Download and uncompress miRDB v6.0 by running:
```
wget https://mirdb.org/download/miRDB_v6.0_prediction_result.txt.gz -P data/
gzip -d miRDB_v6.0_prediction_result.txt.gz
```

The first column of the dataset lists the miRNAs, while the second column lists the mRNAs (more specifically, the RefSeq IDs of the mRNAs).
However, our gene sets list the mRNAs of interest (i.e., those involved in regulated cell death) using their gene symbols. <br>
Hence, we need to perform some preprocessing to convert the gene symbols to RefSeq IDs.

```{r}
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
```

Alternative miRNA-mRNA mapping tools/databases:

- mirTarRnaSeq: https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-022-08558-w
- multiMiR: https://academic.oup.com/nar/article/42/17/e133/2902504

#### Necroptosis

Fetch the necroptosis gene set.

```{r}
necroptosis.genes <- msigdbr(species = "human", category = "C5", subcategory = "GO:BP") %>%
  dplyr::filter(gs_name == "GOBP_NECROPTOTIC_SIGNALING_PATHWAY")
necroptosis.genes
```

Get the gene symbols of the genes included in the necroptosis gene set, and convert them to RefSeq IDs.

```{r}
rcd_gene_symbols <- necroptosis.genes$gene_symbol
rcd_refseq <- getBM(attributes = c("refseq_mrna", "hgnc_symbol"), filters = "hgnc_symbol", values = rcd_gene_symbols, mart = mart)
rcd_refseq
```

Write the RefSeq IDs of the mRNAs of interest to a file.

```{r, echo = TRUE, results="hide"}
rcd_mrna_file <- "temp/necroptosis-genes-refseq.txt"
rcd_refseq.unique_ids <- unique(unlist(rcd_refseq$refseq_mrna))
rcd_refseq.unique_ids <- rcd_refseq.unique_ids[!sapply(rcd_refseq.unique_ids, identical, "")]

# Regenerate the file every time
if (file.exists(rcd_mrna_file)) {
  file.remove(rcd_mrna_file)
}

invisible(lapply(rcd_refseq.unique_ids, write, rcd_mrna_file, append = TRUE, ncolumns = 1))
```

Run the following Python script to fetch the miRNAs targeting the mRNAs of interest:
```
python util/get-mirna.py --mrna-list temp/necroptosis-genes-refseq.txt --output temp/necroptosis-mirna.txt
```

Running this script should generate a file named `necroptosis-mirna.txt` inside the `temp` directory.

Fetch the miRNAs from the said file.

```{r}
necroptosis.mirna <- read.table("temp/necroptosis-mirna.txt")
necroptosis.mirna <- unique(unlist(necroptosis.mirna$V1))
```

Filter the genes to include only those in the necroptosis gene set.

```{r, echo = TRUE, results="hide"}
coad_necroptosis <- count_matrix[rownames(count_matrix) %in% necroptosis.mirna, ]
coad_necroptosis <- coad_necroptosis[, rownames(samples)]

# Check if all samples in the counts dataframe are in the samples dataframe
all(colnames(coad_necroptosis) == rownames(samples))
```

Perform differential miRNA expression analysis.

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = coad_necroptosis,
  colData = samples,
  design = ~type
)
dds <- filter_genes(dds, min_count = 10)
dds$type <- relevel(dds$type, ref = "normal")
dds <- DESeq(dds)
res <- results(dds)
summary(res)
```

Prettify the display of results.

```{r}
deseq.results <- res
deseq.bbl.data <- data.frame(
  row.names = rownames(deseq.results),
  baseMean = deseq.results$baseMean,
  log2FoldChange = deseq.results$log2FoldChange,
  lfcSE = deseq.results$lfcSE,
  stat = deseq.results$stat,
  pvalue = deseq.results$pvalue,
  padj = deseq.results$padj,
  cancer_type = "Colon"
)
deseq.bbl.data
```

Filter based on p-value and log fold change cutoffs.

```{r}
deseq.bbl.data.filtered <- dplyr::filter(deseq.bbl.data, abs(log2FoldChange) >= 1.5 & padj < 0.05)
deseq.bbl.data.filtered
```

Plot the results.

```{r, fig.width=10,fig.height=15}
ggplot(deseq.bbl.data.filtered, aes(x = cancer_type, y = rownames(deseq.bbl.data.filtered), size = padj, fill = log2FoldChange)) +
  geom_point(alpha = 0.5, shape = 21, color = "black") +
  scale_size(trans = "reverse") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", limits = c(min(deseq.bbl.data.filtered$log2FoldChange),max(deseq.bbl.data.filtered$log2FoldChange))) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(legend.position = "bottom") +
  labs(size = "FDR", fill = "log2 FC", x = "Cancer type", y = "miRNA")
```
#### Ferroptosis

Fetch the ferroptosis gene set.

```{r}
ferroptosis.genes <- msigdbr(species = "human", category = "C2", subcategory = "CP:WIKIPATHWAYS") %>%
  dplyr::filter(gs_name == "WP_FERROPTOSIS")
ferroptosis.genes
```

Get the gene symbols of the genes included in the ferroptosis gene set, and convert them to RefSeq IDs.

```{r}
rcd_gene_symbols <- ferroptosis.genes$gene_symbol
rcd_refseq <- getBM(attributes = c("refseq_mrna", "hgnc_symbol"), filters = "hgnc_symbol", values = rcd_gene_symbols, mart = mart)
rcd_refseq
```

Write the RefSeq IDs of the mRNAs of interest to a file.

```{r, echo = TRUE, results="hide"}
rcd_mrna_file <- "temp/ferroptosis-genes-refseq.txt"
rcd_refseq.unique_ids <- unique(unlist(rcd_refseq$refseq_mrna))
rcd_refseq.unique_ids <- rcd_refseq.unique_ids[!sapply(rcd_refseq.unique_ids, identical, "")]

# Regenerate the file every time
if (file.exists(rcd_mrna_file)) {
  file.remove(rcd_mrna_file)
}

invisible(lapply(rcd_refseq.unique_ids, write, rcd_mrna_file, append = TRUE, ncolumns = 1))
```
Run the following Python script to fetch the miRNAs targeting the mRNAs of interest:

```
python util/get-mirna.py --mrna-list temp/ferroptosis-genes-refseq.txt --output temp/ferroptosis-mirna.txt
```

Running this script should generate a file named `ferroptosis-mirna.txt` inside the `temp` directory.

Fetch the miRNAs from the said file.

```{r}
ferroptosis.mirna <- read.table("temp/ferroptosis-mirna.txt")
ferroptosis.mirna <- unique(unlist(ferroptosis.mirna$V1))
```

Filter the genes to include only those in the ferroptosis gene set.

```{r, echo = TRUE, results="hide"}
coad_ferroptosis <- count_matrix[rownames(count_matrix) %in% ferroptosis.mirna, ]
coad_ferroptosis <- coad_ferroptosis[, rownames(samples)]

# Check if all samples in the counts dataframe are in the samples dataframe
all(colnames(coad_ferroptosis) == rownames(samples))
```
Perform differential miRNA expression analysis.

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = coad_ferroptosis,
  colData = samples,
  design = ~type
)
dds <- filter_genes(dds, min_count = 10)
dds$type <- relevel(dds$type, ref = "normal")
dds <- DESeq(dds)
res <- results(dds)
summary(res)
```
Prettify the display of results.

```{r}
deseq.results <- res
deseq.bbl.data <- data.frame(
  row.names = rownames(deseq.results),
  baseMean = deseq.results$baseMean,
  log2FoldChange = deseq.results$log2FoldChange,
  lfcSE = deseq.results$lfcSE,
  stat = deseq.results$stat,
  pvalue = deseq.results$pvalue,
  padj = deseq.results$padj,
  cancer_type = "Colon"
)
deseq.bbl.data
```

Filter based on p-value and log fold change cutoffs.

```{r}
deseq.bbl.data.filtered <- dplyr::filter(deseq.bbl.data, abs(log2FoldChange) >= 1.5 & padj < 0.05)
deseq.bbl.data.filtered
```

Plot the results.

```{r, fig.width=10,fig.height=45}
ggplot(deseq.bbl.data.filtered, aes(x = cancer_type, y = rownames(deseq.bbl.data.filtered), size = padj, fill = log2FoldChange)) +
  geom_point(alpha = 0.5, shape = 21, color = "black") +
  scale_size(trans = "reverse") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", limits = c(min(deseq.bbl.data.filtered$log2FoldChange),max(deseq.bbl.data.filtered$log2FoldChange))) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(legend.position = "bottom") +
  labs(size = "FDR", fill = "log2 FC", x = "Cancer type", y = "miRNA")
```

#### Pyroptosis

Fetch the pyroptosis gene set.

```{r}
pyroptosis.genes <- msigdbr(species = "human", category = "C2", subcategory = "CP:REACTOME") %>%
  dplyr::filter(gs_name == "REACTOME_PYROPTOSIS")
pyroptosis.genes
```

Get the gene symbols of the genes included in the pyroptosis gene set, and convert them to RefSeq IDs.

```{r}
rcd_gene_symbols <- pyroptosis.genes$gene_symbol
rcd_refseq <- getBM(attributes = c("refseq_mrna", "hgnc_symbol"), filters = "hgnc_symbol", values = rcd_gene_symbols, mart = mart)
rcd_refseq
```
Write the RefSeq IDs of the mRNAs of interest to a file.

```{r, echo = TRUE, results="hide"}
rcd_mrna_file <- "temp/pyroptosis-genes-refseq.txt"
rcd_refseq.unique_ids <- unique(unlist(rcd_refseq$refseq_mrna))
rcd_refseq.unique_ids <- rcd_refseq.unique_ids[!sapply(rcd_refseq.unique_ids, identical, "")]

# Regenerate the file every time
if (file.exists(rcd_mrna_file)) {
  file.remove(rcd_mrna_file)
}

invisible(lapply(rcd_refseq.unique_ids, write, rcd_mrna_file, append = TRUE, ncolumns = 1))
```

Run the following Python script to fetch the miRNAs targeting the mRNAs of interest:

```
python util/get-mirna.py --mrna-list temp/pyroptosis-genes-refseq.txt --output temp/pyroptosis-mirna.txt
```

Running this script should generate a file named `pyroptosis-mirna.txt` inside the `temp` directory.

Fetch the miRNAs from the said file.

```{r}
pyroptosis.mirna <- read.table("temp/pyroptosis-mirna.txt")
pyroptosis.mirna <- unique(unlist(pyroptosis.mirna$V1))
```

Filter the genes to include only those in the pyroptosis gene set.

```{r, echo = TRUE, results="hide"}
coad_pyroptosis <- count_matrix[rownames(count_matrix) %in% pyroptosis.mirna, ]
coad_pyroptosis <- coad_pyroptosis[, rownames(samples)]

# Check if all samples in the counts dataframe are in the samples dataframe
all(colnames(coad_pyroptosis) == rownames(samples))
```

Perform differential miRNA expression analysis.

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = coad_pyroptosis,
  colData = samples,
  design = ~type
)
dds <- filter_genes(dds, min_count = 10)
dds$type <- relevel(dds$type, ref = "normal")
dds <- DESeq(dds)
res <- results(dds)
summary(res)
```
Prettify the display of results.

```{r}
deseq.results <- res
deseq.bbl.data <- data.frame(
  row.names = rownames(deseq.results),
  baseMean = deseq.results$baseMean,
  log2FoldChange = deseq.results$log2FoldChange,
  lfcSE = deseq.results$lfcSE,
  stat = deseq.results$stat,
  pvalue = deseq.results$pvalue,
  padj = deseq.results$padj,
  cancer_type = "Colon"
)
deseq.bbl.data
```

Filter based on p-value and log fold change cutoffs.

```{r}
deseq.bbl.data.filtered <- dplyr::filter(deseq.bbl.data, abs(log2FoldChange) >= 1.5 & padj < 0.05)
deseq.bbl.data.filtered
```


Plot the results.

```{r, fig.width=10,fig.height=30}
ggplot(deseq.bbl.data.filtered, aes(x = cancer_type, y = rownames(deseq.bbl.data.filtered), size = padj, fill = log2FoldChange)) +
  geom_point(alpha = 0.5, shape = 21, color = "black") +
  scale_size(trans = "reverse") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", limits = c(min(deseq.bbl.data.filtered$log2FoldChange),max(deseq.bbl.data.filtered$log2FoldChange))) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(legend.position = "bottom") +
  labs(size = "FDR", fill = "log2 FC", x = "Cancer type", y = "miRNA")
```