---
title: "Differential Gene Expression Analysis"
subtitle: "Pan-cancer | Multiple RCD types"
author: 
  - Mark Edward M. Gonzales^[De La Salle University, Manila, Philippines, gonzales.markedward@gmail.com]
  - Dr. Anish M.S. Shrestha^[De La Salle University, Manila, Philippines, anish.shrestha@dlsu.edu.ph]
output: html_notebook
---

## I. Preliminaries

### Loading libraries

```{r, warning=FALSE, message=FALSE}
library("tidyverse")
library("tibble")
library("msigdbr")
library("ggplot2")
library("TCGAbiolinks")
library("RNAseqQC")
library("DESeq2")
library("ensembldb")
library("purrr")
library("magrittr")
library("vsn")
library("matrixStats")
library("dplyr")
library("grex")
```

## II. Downloading the TCGA gene expression data 

Create a function for downloading TCGA gene expression data. 

For more detailed documentation, refer to `2. Differential Gene Expression Analysis - TCGA.rmd`.

```{r}
query_and_filter_samples <- function(project) {
  query_tumor <- GDCquery(
    project = project,
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    experimental.strategy = "RNA-Seq",
    workflow.type = "STAR - Counts",
    access = "open",
    sample.type = "Primary Tumor"
  )
  tumor <- getResults(query_tumor)
  
  query_normal <- GDCquery(
    project = project,
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    experimental.strategy = "RNA-Seq",
    workflow.type = "STAR - Counts",
    access = "open",
    sample.type = "Solid Tissue Normal"
  )
  normal <- getResults(query_normal)
  
  submitter_ids <- inner_join(tumor, normal, by = "cases.submitter_id") %>%
    dplyr::select(cases.submitter_id)
  tumor <- tumor %>%
    dplyr::filter(cases.submitter_id %in% submitter_ids$cases.submitter_id)
  normal <- normal %>%
    dplyr::filter(cases.submitter_id %in% submitter_ids$cases.submitter_id)
  
  samples <- rbind(tumor, normal)
  unique(samples$sample_type)
  
  query_project <- GDCquery(
    project = project,
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    experimental.strategy = "RNA-Seq",
    workflow.type = "STAR - Counts",
    access = "open",
    sample.type = c("Solid Tissue Normal", "Primary Tumor"),
    barcode = as.list(samples$sample.submitter_id)
  )
  
  # If this is your first time running this notebook (i.e., you have not yet downloaded the results of the query in the previous block), 
  # uncomment the line below
  
  # GDCdownload(query_project)
  
  return (list(samples = samples, query_project = query_project))
}
```

Download the TCGA gene expression data for different cancer types.

Refer to this link for the list of TCGA cancer type abbreviations: https://gdc.cancer.gov/resources-tcga-users/tcga-code-tables/tcga-study-abbreviations

```{r, message=FALSE}
projects <- c("TCGA-LUSC", "TCGA-COAD", "TCGA-KICH", "TCGA-KIRC", "TCGA-PRAD",
              "TCGA-BRCA", "TCGA-HNSC", "TCGA-KIRP", "TCGA-LIHC", "TCGA-STAD",
              "TCGA-THCA", "TCGA-BLCA", "TCGA-LUAD", "TCGA-ESCA")

samples <- list()
project_data <- list()

for (project in projects) {
  result <- query_and_filter_samples(project)
  
  samples[[project]] <- result$samples
  project_data[[project]] <- result$query_project
}
```


Running the code block above should generate and populate a directory named `GDCdata`.

## III. Data preprocessing

Construct the RNA-seq count matrix for each cancer type.

```{r, echo = TRUE, message = FALSE, results="hide"}
tcga_data <- list()
tcga_matrix <- list()

for (project in projects) {
  tcga_data[[project]] <- GDCprepare(project_data[[project]], summarizedExperiment = TRUE)
}
```

```{r}
for (project in projects) {
  count_matrix <- assay(tcga_data[[project]], "unstranded")
  
  # Remove duplicate entries
  count_matrix_df <- data.frame(count_matrix)
  count_matrix_df <- count_matrix_df[!duplicated(count_matrix_df), ]
  count_matrix <- data.matrix(count_matrix_df)
  rownames(count_matrix) <- cleanid(rownames(count_matrix))
  count_matrix <- count_matrix[!(duplicated(rownames(count_matrix)) | duplicated(rownames(count_matrix), fromLast = TRUE)), ]
  
  tcga_matrix[[project]] <- count_matrix
}
```
Format the `samples` table so that it can be fed as input to DESeq2.

```{r}
for (project in projects) {
  rownames(samples[[project]]) <- samples[[project]]$cases
  samples[[project]] <- samples[[project]] %>%
    dplyr::select(case = "cases.submitter_id", type = "sample_type")
  samples[[project]]$type <- str_replace(samples[[project]]$type, "Solid Tissue Normal", "normal")
  samples[[project]]$type <- str_replace(samples[[project]]$type, "Primary Tumor", "tumor")
}
```

DESeq2 requires the row names of `samples` should be identical to the column names of `count_matrix`.

```{r, echo = TRUE, results="hide"}
for (project in projects) {
  colnames(tcga_matrix[[project]]) <- gsub(x = colnames(tcga_matrix[[project]]), pattern = "\\.", replacement = "-")
  tcga_matrix[[project]] <- tcga_matrix[[project]][, rownames(samples[[project]])]
  
  # Sanity check
  print(all(colnames(tcga_matrix[[project]]) == rownames(samples[[project]])))
}
```

## IV. Differential gene expression analysis

References: 

- Official documentation: https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html
- Good balance of theory and hands-on: https://hbctraining.github.io/DGE_workshop/lessons/04_DGE_DESeq2_analysis.html
- Quality control: https://cran.r-project.org/web/packages/RNAseqQC/vignettes/introduction.html

Construct the `DESeqDataSet` object for each cancer type.

```{r}
dds_results <- list()

for (project in projects) {
  dds_results[[project]] <- DESeqDataSetFromMatrix(
    countData = tcga_matrix[[project]],
    colData = samples[[project]],
    design = ~type
  )
}
```


### Regulated Cell Death

We obtain the gene sets from RCDdb: https://pmc.ncbi.nlm.nih.gov/articles/PMC11384979/

Download the gene sets by running:

```
wget http://chenyclab.com/RCDdb/download/Apoptosis.csv -P data/RCDdb/
wget http://chenyclab.com/RCDdb/download/Entotic_cell_death.csv -P data/RCDdb/
wget http://chenyclab.com/RCDdb/download/Necroptosis.csv -P data/RCDdb/
wget http://chenyclab.com/RCDdb/download/Alkaliptosis.csv -P data/RCDdb/
wget http://chenyclab.com/RCDdb/download/Ferroptosis.csv -P data/RCDdb/
wget http://chenyclab.com/RCDdb/download/NETotic_cell_death.csv -P data/RCDdb/
wget http://chenyclab.com/RCDdb/download/Autophagy_dependent_cell_death.csv -P data/RCDdb/
wget http://chenyclab.com/RCDdb/download/Immunogenic_cell_death.csv -P data/RCDdb/
wget http://chenyclab.com/RCDdb/download/Oxeiptosis.csv -P data/RCDdb/
wget http://chenyclab.com/RCDdb/download/Cuproptosis.csv -P data/RCDdb/
wget http://chenyclab.com/RCDdb/download/Lysosome_dependent_cell_death.csv -P data/RCDdb/
wget http://chenyclab.com/RCDdb/download/Parthanatos.csv -P data/RCDdb/
wget http://chenyclab.com/RCDdb/download/Disulfidptosis.csv -P data/RCDdb/
wget http://chenyclab.com/RCDdb/download/MPT_driven_necrosis.csv -P data/RCDdb/
wget http://chenyclab.com/RCDdb/download/Pyroptosis.csv -P data/RCDdb/
```

```{r}
RCDdb <- "data/RCDdb/"
```

Write utility functions for filtering the gene sets, performing differential gene expression analysis, and plotting the results.

```{r}
filter_gene_set_and_perform_dgea <- function(genes) {
  tcga_rcd <- list()

  for (project in projects) {
    rownames(genes) <- genes$gene_id
    tcga_rcd[[project]] <- tcga_matrix[[project]][rownames(tcga_matrix[[project]]) %in% genes$gene_id, ]
    tcga_rcd[[project]] <- tcga_rcd[[project]][, rownames(samples[[project]])]
  }
  
  dds_rcd <- list()
  res_rcd <- list()
  
  for (project in projects) {
    print(project)
    print("=============")
    dds <- DESeqDataSetFromMatrix(
      countData = tcga_rcd[[project]],
      colData = samples[[project]],
      design = ~type
    )
    dds <- filter_genes(dds, min_count = 10)
    dds$type <- relevel(dds$type, ref = "normal")
    dds_rcd[[project]] <- DESeq(dds)
    res_rcd[[project]] <- results(dds_rcd[[project]])
  }
  
  deseq.bbl.data <- list()
  
  for (project in projects) {
    deseq.results <- res_rcd[[project]]
    deseq.bbl.data[[project]] <- data.frame(
      row.names = rownames(deseq.results),
      baseMean = deseq.results$baseMean,
      log2FoldChange = deseq.results$log2FoldChange,
      lfcSE = deseq.results$lfcSE,
      stat = deseq.results$stat,
      pvalue = deseq.results$pvalue,
      padj = deseq.results$padj,
      cancer_type = project,
      gene_symbol = genes[rownames(deseq.results), "gene"]
    )
  }
  
  deseq.bbl.data.combined <- bind_rows(deseq.bbl.data)
  deseq.bbl.data.combined <- dplyr::filter(deseq.bbl.data.combined, abs(log2FoldChange) >= 1.5 & padj < 0.05)
  
  return (deseq.bbl.data.combined)
}
```

```{r}
plot_dgea <- function(deseq.bbl.data.combined) {
  sizes <- c("<10^-15" = 4, "10^-10" = 3, "10^-5" = 2, "0.05" = 1)
  
  deseq.bbl.data.combined <- deseq.bbl.data.combined %>%
  mutate(fdr_category = cut(padj,
                            breaks = c(-Inf, 1e-15, 1e-10, 1e-5, 0.05),
                            labels = c("<10^-15", "10^-10", "10^-5", "0.05"),
                            right = FALSE))

  top_genes <- deseq.bbl.data.combined %>%
    group_by(cancer_type) %>%
    mutate(rank = rank(-abs(log2FoldChange))) %>%
    dplyr::filter(rank <= 10) %>%
    ungroup()
  
  ggplot(top_genes, aes(y=cancer_type, x=gene_symbol, size=fdr_category, fill=log2FoldChange)) +
      geom_point(alpha=0.5, shape=21, color="black") +
      scale_size_manual(values = sizes) +
      scale_fill_gradient2(low = "blue", mid = "white", high = "red", limits = c(min(deseq.bbl.data.combined$log2FoldChange),max(deseq.bbl.data.combined$log2FoldChange))) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(size = 9, angle = 90, hjust = 1)
      ) +
      theme(legend.position="bottom") +
      theme(legend.position = "bottom")+
      labs(size = "Adjusted p-value", fill = "log2 FC", y = "Cancer type", x = "Gene")
}
```

#### 1. Necroptosis

Fetch the gene set of interest.

```{r}
genes <- read.csv(paste0(RCDdb, "Necroptosis.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```
Filter the genes to include only those in the gene set of interest, and then perform differential gene expression analysis.

```{r}
deseq.bbl.data.combined <- filter_gene_set_and_perform_dgea(genes)
deseq.bbl.data.combined
```

Plot the results.

```{r, fig.width = 15, fig.height=5}
plot_dgea(deseq.bbl.data.combined)
```

#### 2. Ferroptosis

Fetch the gene set of interest.

```{r}
genes <- read.csv(paste0(RCDdb, "Ferroptosis.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Filter the genes to include only those in the gene set of interest, and then perform differential gene expression analysis.

```{r}
deseq.bbl.data.combined <- filter_gene_set_and_perform_dgea(genes)
deseq.bbl.data.combined
```

Plot the results.

```{r, fig.width = 15, fig.height=5}
plot_dgea(deseq.bbl.data.combined)
```

#### 3. Pyroptosis

Fetch the gene set of interest.

```{r}
genes <- read.csv(paste0(RCDdb, "Pyroptosis.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Filter the genes to include only those in the gene set of interest, and then perform differential gene expression analysis.

```{r}
deseq.bbl.data.combined <- filter_gene_set_and_perform_dgea(genes)
deseq.bbl.data.combined
```

Plot the results.

```{r, fig.width = 15, fig.height=5}
plot_dgea(deseq.bbl.data.combined)
```

#### 4. Alkaliptosis

Fetch the gene set of interest.

```{r}
genes <- read.csv(paste0(RCDdb, "Alkaliptosis.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Filter the genes to include only those in the gene set of interest, and then perform differential gene expression analysis.

```{r}
deseq.bbl.data.combined <- filter_gene_set_and_perform_dgea(genes)
deseq.bbl.data.combined
```

Plot the results.

```{r, fig.width = 15, fig.height=5}
plot_dgea(deseq.bbl.data.combined)
```

#### 5. Apoptosis

Fetch the gene set of interest.

```{r}
genes <- read.csv(paste0(RCDdb, "Apoptosis.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Filter the genes to include only those in the gene set of interest, and then perform differential gene expression analysis.

```{r}
deseq.bbl.data.combined <- filter_gene_set_and_perform_dgea(genes)
deseq.bbl.data.combined
```

Plot the results.

```{r, fig.width = 15, fig.height=5}
plot_dgea(deseq.bbl.data.combined)
```

#### 6. Autophagy-Dependent Cell Death

Fetch the gene set of interest.

```{r}
genes <- read.csv(paste0(RCDdb, "Autophagy_dependent_cell_death.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Filter the genes to include only those in the gene set of interest, and then perform differential gene expression analysis.

```{r}
deseq.bbl.data.combined <- filter_gene_set_and_perform_dgea(genes)
deseq.bbl.data.combined
```

Plot the results.

```{r, fig.width = 15, fig.height=5}
plot_dgea(deseq.bbl.data.combined)
```

#### 7. Cuproptosis

Fetch the gene set of interest.

```{r}
genes <- read.csv(paste0(RCDdb, "Cuproptosis.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Filter the genes to include only those in the gene set of interest, and then perform differential gene expression analysis.

```{r}
deseq.bbl.data.combined <- filter_gene_set_and_perform_dgea(genes)
deseq.bbl.data.combined
```

Plot the results.

```{r, fig.width = 15, fig.height=5}
plot_dgea(deseq.bbl.data.combined)
```

#### 8. Disulfidptosis

Fetch the gene set of interest.

```{r}
genes <- read.csv(paste0(RCDdb, "Disulfidptosis.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Filter the genes to include only those in the gene set of interest, and then perform differential gene expression analysis.

```{r}
deseq.bbl.data.combined <- filter_gene_set_and_perform_dgea(genes)
deseq.bbl.data.combined
```

Plot the results.

```{r, fig.width = 15, fig.height=5}
plot_dgea(deseq.bbl.data.combined)
```

#### 9. Entotic Cell Death

Fetch the gene set of interest.

```{r}
genes <- read.csv(paste0(RCDdb, "Entotic_cell_death.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Filter the genes to include only those in the gene set of interest, and then perform differential gene expression analysis.

```{r}
deseq.bbl.data.combined <- filter_gene_set_and_perform_dgea(genes)
deseq.bbl.data.combined
```

Plot the results.

```{r, fig.width = 15, fig.height=5}
plot_dgea(deseq.bbl.data.combined)
```

#### 10. Immunogenic Cell Death

Fetch the gene set of interest.

```{r}
genes <- read.csv(paste0(RCDdb, "Immunogenic_cell_death.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Filter the genes to include only those in the gene set of interest, and then perform differential gene expression analysis.

```{r}
deseq.bbl.data.combined <- filter_gene_set_and_perform_dgea(genes)
deseq.bbl.data.combined
```

Plot the results.

```{r, fig.width = 15, fig.height=5}
plot_dgea(deseq.bbl.data.combined)
```

#### 11. Lysosome-Dependent Cell Death

Fetch the gene set of interest.

```{r}
genes <- read.csv(paste0(RCDdb, "Lysosome_dependent_cell_death.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Filter the genes to include only those in the gene set of interest, and then perform differential gene expression analysis.

```{r}
deseq.bbl.data.combined <- filter_gene_set_and_perform_dgea(genes)
deseq.bbl.data.combined
```

Plot the results.

```{r, fig.width = 15, fig.height=5}
plot_dgea(deseq.bbl.data.combined)
```

#### 12. MPT-Driven Necrosis

Fetch the gene set of interest.

```{r}
genes <- read.csv(paste0(RCDdb, "MPT_driven_necrosis.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Filter the genes to include only those in the gene set of interest, and then perform differential gene expression analysis.

```{r}
deseq.bbl.data.combined <- filter_gene_set_and_perform_dgea(genes)
deseq.bbl.data.combined
```

Plot the results.

```{r, fig.width = 15, fig.height=5}
plot_dgea(deseq.bbl.data.combined)
```

#### 13. NETotic Cell Death

Fetch the gene set of interest.

```{r}
genes <- read.csv(paste0(RCDdb, "NETotic_cell_death.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Filter the genes to include only those in the gene set of interest, and then perform differential gene expression analysis.

```{r}
deseq.bbl.data.combined <- filter_gene_set_and_perform_dgea(genes)
deseq.bbl.data.combined
```

Plot the results.

```{r, fig.width = 15, fig.height=5}
plot_dgea(deseq.bbl.data.combined)
```

#### 14. Oxeiptosis

Fetch the gene set of interest.

```{r}
genes <- read.csv(paste0(RCDdb, "Oxeiptosis.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Filter the genes to include only those in the gene set of interest, and then perform differential gene expression analysis.

```{r}
deseq.bbl.data.combined <- filter_gene_set_and_perform_dgea(genes)
deseq.bbl.data.combined
```

Plot the results.

```{r, fig.width = 15, fig.height=5}
plot_dgea(deseq.bbl.data.combined)
```


#### 15. Parthanatos

Fetch the gene set of interest.

```{r}
genes <- read.csv(paste0(RCDdb, "Parthanatos.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Filter the genes to include only those in the gene set of interest, and then perform differential gene expression analysis.

```{r}
deseq.bbl.data.combined <- filter_gene_set_and_perform_dgea(genes)
deseq.bbl.data.combined
```

Plot the results.

```{r, fig.width = 15, fig.height=5}
plot_dgea(deseq.bbl.data.combined)
```