---
title: "Gene Expression Analysis"
subtitle: "MCF7 Breast Cancer Cell Line | GSE288929 | Necroptosis, Ferroptosis & Pyroptosis"
author: 
  - Mark Edward M. Gonzales^[De La Salle University, Manila, Philippines, gonzales.markedward@gmail.com]
  - Dr. Anish M.S. Shrestha^[De La Salle University, Manila, Philippines, anish.shrestha@dlsu.edu.ph]
output: html_notebook
---

## I. Preliminaries

### Loading libraries

```{r, warning=FALSE, message=FALSE}
library("tidyverse")
library("tibble")
library("msigdbr")
library("ggplot2")
library("ensembldb")
library("purrr")
library("magrittr")
library("matrixStats")
library("dplyr")
library("grex")
library("gplots")
library("RColorBrewer")
```

### Constants
```{r}
DATA_DIR <- "../data/public/GEO/MCF7/"
```

## Loading the Expression Data

The expression data are taken from this study: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE288929

The relevant columns for MCF7 are `TC00000367_D05` to `TC00000367_D07`, as stated here:
- https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM8779050
- https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM8779051
- https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM8779052

Each column refers to one replicate.

```{r}
mcf7.expression <- read.delim(paste0(DATA_DIR, "GSE288929_httr_atlas_counts.csv"), as.is = TRUE, header = TRUE, row.names = 1, sep=",")
mcf7.expression <- rownames_to_column(mcf7.expression, "Probe_ID") %>% rename("Probe_ID" = "ID_REF")
mcf7.expression <- mcf7.expression[c("ID_REF", "TC00000367_D05", "TC00000367_D06", "TC00000367_D07")]
mcf7.expression
```

Map the probe IDs to Ensembl accessions.

```{r}
probe2ensembl <- read.delim(paste0(DATA_DIR, "GSE288929_httr_atlas_probes.csv"), as.is = TRUE, header = TRUE, row.names = 1, sep=",")
probe2ensembl <- probe2ensembl[c("Probe_Name", "Ensembl_Gene")] %>% 
  rename("Probe_Name" = "ID_REF", "Ensembl_Gene" = "gene_id")
rownames(probe2ensembl) <- NULL
probe2ensembl
```

```{r}
mcf7.expression <- left_join(mcf7.expression, probe2ensembl)
mcf7.expression
```

## Exploratory Data Analysis

We load the gene sets from RCDdb: https://pubmed.ncbi.nlm.nih.gov/39257527/

```{r}
RCDdb <- "../data/public/rcd-gene-list/RCDdb/"
```

### Necroptosis

Load the gene set.

```{r}
genes <- read.csv(paste0(RCDdb, "Necroptosis.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Get the normalized expression data for the genes in the gene set.

```{r}
tpm.df <- mcf7.expression %>% dplyr::filter(gene_id %in% genes$gene_id)
tpm.df <- left_join(tpm.df, genes %>% dplyr::select(gene_id, gene), by = c("gene_id" = "gene_id"))
tpm.df <- distinct(tpm.df, gene, .keep_all = TRUE)
rownames(tpm.df) <- tpm.df$gene
tpm.df <- subset(tpm.df, select = -c(gene_id, ID_REF, gene))
tpm.df <- tpm.df[order(row.names(tpm.df)), , drop = FALSE]
tpm.df
```

Plot the results.

```{r, fig.height=30, fig.width=10}
tpm.matrix <- as.matrix(tpm.df)
heatmap.2(tpm.matrix,
  srtCol = 360,
  cellnote = tpm.matrix,
  dendrogram = "none", Colv = FALSE, Rowv = FALSE,
  col = brewer.pal(n = 9, name = "BuPu")[5:9], trace = "none", key = FALSE, lwid = c(0.1, 4), lhei = c(0.1, 4),
  cexCol = 1, cexRow = 0.75, symm = TRUE
)
```
### Ferroptosis

Load the gene set.

```{r}
genes <- read.csv(paste0(RCDdb, "Ferroptosis.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Get the normalized expression data for the genes in the gene set.

```{r}
tpm.df <- mcf7.expression %>% dplyr::filter(gene_id %in% genes$gene_id)
tpm.df <- left_join(tpm.df, genes %>% dplyr::select(gene_id, gene), by = c("gene_id" = "gene_id"))
tpm.df <- distinct(tpm.df, gene, .keep_all = TRUE)
rownames(tpm.df) <- tpm.df$gene
tpm.df <- subset(tpm.df, select = -c(gene_id, ID_REF, gene))
tpm.df <- tpm.df[order(row.names(tpm.df)), , drop = FALSE]
tpm.df
```

Plot the results.

```{r, fig.height=150, fig.width=10}
tpm.matrix <- as.matrix(tpm.df)
heatmap.2(tpm.matrix,
  srtCol = 360,
  cellnote = tpm.matrix,
  dendrogram = "none", Colv = FALSE, Rowv = FALSE,
  col = brewer.pal(n = 9, name = "BuPu")[5:9], trace = "none", key = FALSE, lwid = c(0.1, 4), lhei = c(0.1, 4),
  cexCol = 1, cexRow = 0.75, symm = TRUE
)
```

### Pyroptosis

Load the gene set.

```{r}
genes <- read.csv(paste0(RCDdb, "Pyroptosis.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Get the normalized expression data for the genes in the gene set.

```{r}
tpm.df <- mcf7.expression %>% dplyr::filter(gene_id %in% genes$gene_id)
tpm.df <- left_join(tpm.df, genes %>% dplyr::select(gene_id, gene), by = c("gene_id" = "gene_id"))
tpm.df <- distinct(tpm.df, gene, .keep_all = TRUE)
rownames(tpm.df) <- tpm.df$gene
tpm.df <- subset(tpm.df, select = -c(gene_id, ID_REF, gene))
tpm.df <- tpm.df[order(row.names(tpm.df)), , drop = FALSE]
tpm.df
```

Plot the results.

```{r, fig.height=20, fig.width=10}
tpm.matrix <- as.matrix(tpm.df)
heatmap.2(tpm.matrix,
  srtCol = 360,
  cellnote = tpm.matrix,
  dendrogram = "none", Colv = FALSE, Rowv = FALSE,
  col = brewer.pal(n = 9, name = "BuPu")[5:9], trace = "none", key = FALSE, lwid = c(0.1, 4), lhei = c(0.1, 4),
  cexCol = 1, cexRow = 0.75, symm = TRUE
)
```