---
title: "Gene Expression Analysis"
subtitle: "Fetal colon cell line FHC | Necroptosis, Ferroptosis & Pyroptosis"
author: 
  - Mark Edward M. Gonzales^[De La Salle University, Manila, Philippines, gonzales.markedward@gmail.com]
  - Dr. Anish M.S. Shrestha^[De La Salle University, Manila, Philippines, anish.shrestha@dlsu.edu.ph]
output: html_notebook
---

## I. Preliminaries

### Loading libraries

```{r, warning=FALSE, message=FALSE}
library("tidyverse")
library("tibble")
library("msigdbr")
library("ggplot2")
library("ensembldb")
library("purrr")
library("magrittr")
library("matrixStats")
library("dplyr")
library("grex")
library("gplots")
library("RColorBrewer")
```

### Constants
```{r}
DATA_DIR <- "../data/public/GEO/FHC/"
```

## Loading the Expression Data

The expression data are taken from this study: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE232211

```{r}
fhc.expression <- read.delim(paste0(DATA_DIR, "GSE232211_norm_counts_TPM_GRCh38.p13_NCBI.tsv"), as.is = TRUE, header = TRUE, row.names = 1)
fhc.expression <- rownames_to_column(fhc.expression, "gene_id")
fhc.expression
```
```{r}
mapping_table <- read.delim(paste0(DATA_DIR, "Human.GRCh38.p13.annot.tsv"), as.is = TRUE, header = TRUE, row.names = 1)
mapping_table <- rownames_to_column(mapping_table, "gene_id")
mapping_table <- subset(mapping_table, EnsemblGeneID != "")
mapping_table
```
Since the gene set for the RCD-related regulators (taken from RCDdb) uses Ensembl Gene IDs, we need to map the accessions.

```{r}
fhc.expression.mapped <-  right_join(fhc.expression, mapping_table, by = join_by(gene_id == gene_id)) %>% distinct(EnsemblGeneID, .keep_all = TRUE)
rownames(fhc.expression.mapped) <- fhc.expression.mapped$EnsemblGeneID
fhc.expression.mapped = fhc.expression.mapped[,!(names(fhc.expression.mapped) %in% colnames(mapping_table))]
fhc.expression.mapped <- rownames_to_column(fhc.expression.mapped, "gene_id")
fhc.expression.mapped
```

## Exploratory Data Analysis

We load the gene sets from RCDdb: https://pubmed.ncbi.nlm.nih.gov/39257527/

```{r}
RCDdb <- "../data/public/rcd-gene-list/RCDdb/"
```

### Necroptosis

Load the gene set.

```{r}
genes <- read.csv(paste0(RCDdb, "Necroptosis.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Get the TPM for the genes in the gene set.

```{r}
tpm.df <- fhc.expression.mapped %>% dplyr::filter(gene_id %in% genes$gene_id)
tpm.df <- left_join(tpm.df, genes %>% dplyr::select(gene_id, gene), by = c("gene_id" = "gene_id"))
rownames(tpm.df) <- tpm.df$gene
tpm.df <- tpm.df %>% dplyr::select(c(1:4))
tpm.df <- subset(tpm.df, select = -c(gene_id) )
tpm.df <- tpm.df[ order(row.names(tpm.df)), ]
tpm.df
```

Plot the results.

```{r, fig.height=30, fig.width=10}
tpm.matrix <- as.matrix(tpm.df)
heatmap.2(tpm.matrix, srtCol=360, cellnote = tpm.matrix, dendrogram="none", Colv=FALSE, Rowv=FALSE,
          col=brewer.pal(n = 9, name = "BuPu")[5:9], trace="none", key = FALSE, lwid=c(0.1,4), lhei=c(0.1,4),
          cexCol=1, cexRow=0.75, symm = TRUE)
```
### Ferroptosis

Load the gene set.

```{r}
genes <- read.csv(paste0(RCDdb, "Ferroptosis.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Get the TPM for the genes in the gene set.

```{r}
tpm.df <- fhc.expression.mapped %>% dplyr::filter(gene_id %in% genes$gene_id)
tpm.df <- left_join(tpm.df, genes %>% dplyr::select(gene_id, gene), by = c("gene_id" = "gene_id"))
rownames(tpm.df) <- tpm.df$gene
tpm.df <- tpm.df %>% dplyr::select(c(1:4))
tpm.df <- subset(tpm.df, select = -c(gene_id) )
tpm.df <- tpm.df[ order(row.names(tpm.df)), ]
tpm.df
```

Plot the results.

```{r, fig.height=150, fig.width=10}
tpm.matrix <- as.matrix(tpm.df)
heatmap.2(tpm.matrix, srtCol=360, cellnote = tpm.matrix, dendrogram="none", Colv=FALSE, Rowv=FALSE,
          col=brewer.pal(n = 9, name = "BuPu")[5:9], trace="none", key = FALSE, lwid=c(0.1,4), lhei=c(0.1,4),
          cexCol=1, cexRow=0.75, symm = TRUE)
```

### Pyroptosis

Load the gene set.

```{r}
genes <- read.csv(paste0(RCDdb, "Pyroptosis.csv"))
genes$gene_id <- cleanid(genes$gene_id)
genes <- distinct(genes, gene_id, .keep_all = TRUE)
genes <- subset(genes, gene_id != "")
genes
```

Get the TPM for the genes in the gene set.

```{r}
tpm.df <- fhc.expression.mapped %>% dplyr::filter(gene_id %in% genes$gene_id)
tpm.df <- left_join(tpm.df, genes %>% dplyr::select(gene_id, gene), by = c("gene_id" = "gene_id"))
rownames(tpm.df) <- tpm.df$gene
tpm.df <- tpm.df %>% dplyr::select(c(1:4))
tpm.df <- subset(tpm.df, select = -c(gene_id) )
tpm.df <- tpm.df[ order(row.names(tpm.df)), ]
tpm.df
```

Plot the results.

```{r, fig.height=20, fig.width=10}
tpm.matrix <- as.matrix(tpm.df)
heatmap.2(tpm.matrix, srtCol=360, cellnote = tpm.matrix, dendrogram="none", Colv=FALSE, Rowv=FALSE,
          col=brewer.pal(n = 9, name = "BuPu")[5:9], trace="none", key = FALSE, lwid=c(0.1,4), lhei=c(0.1,4),
          cexCol=1, cexRow=0.75, symm = TRUE)
```